---
title: "Random sampling from groups"
description: |
  How to draw random samples from several populations (that might have different distributions)
author:
  - name: Ken Butler
    url: http://ritsokiguess.site/blog
date: 2022-05-09
output:
  distill::distill_article:
    self_contained: false
---

# Packages

```{r}
library(tidyverse)
```


# Introduction

In a previous post, I discussed how we might sample in groups, where each group was a sample from a different population. I introduced this function:

```{r}
gen_sample <- function(n, mean, sd) {
  tibble(gp = c("x", "y"), n = n, mean = mean, sd = sd) %>% 
    rowwise() %>% 
    mutate(z = list(rnorm(n, mean, sd))) %>% 
    unnest(z) %>% 
    select(gp, z)
}
```

that samples from normal populations with possibly different means, SDs, and sample sizes in different groups.

# Explanation

This is (more or less) the same explanation that appeared at the end of the previous post, so feel free to skip if you have read it before.

The first step is to make a data frame with one row for each sample that will be generated. This uses the inputs to the function above, so we will make some up:

```{r}
n <- c(5, 3)
mean <- c(20, 10)
sd <- c(2, 1)
tibble(gp = c("x", "y"), n = n, mean = mean, sd = sd) 
```

Evidently, in a function for public consumption, you would check that all the inputs are the same length, or you would rely on `tibble` telling you that only vectors of length 1 are recycled.^[So, for example, if both your sample sizes are the same, you could define eg `n <- 10` and it would get expanded to length 2 in the function.] The groups are for no good reason called `x` and `y`.

The next two lines generate random samples, one for each group, according to the specifications, and store them each in one cell of the two-row spreadsheet:

```{r}
  tibble(gp = c("x", "y"), n = n, mean = mean, sd = sd) %>% 
    rowwise() %>% 
    mutate(z = list(rnorm(n, mean, sd))) 
```

The new column `z` is a list column, since the top cell of the column is a vector of length 5, and the bottom cell is a vector of length 3. To actually see the values they contain, we `unnest` `z`:

```{r}
  tibble(gp = c("x", "y"), n = n, mean = mean, sd = sd) %>% 
    rowwise() %>% 
    mutate(z = list(rnorm(n, mean, sd))) %>% 
    unnest(z)
```

and, finally, the middle three columns were only used to generate the values in `z`, so they can be thrown away now by `select`ing only `gp` and `z`.

The `rowwise` is necessary:

```{r}
  tibble(gp = c("x", "y"), n = n, mean = mean, sd = sd) %>% 
    mutate(z = list(rnorm(n, mean, sd))) %>% 
    unnest(z)
```

because `rnorm` is vectorized, and for the `x` sample, R will draw one sampled value from each normal distribution, and then repeat the same values for the `y` sample. This is very much *not* what we want.

The same idea can be used to draw random chi-squared data (say):

```{r}
tibble(df = c(2,6)) %>% 
  rowwise() %>% 
  mutate(z = list(rchisq(5, df))) %>% 
  unnest(z)
```

(five values from $\chi^2_2$, followed by five from $\chi^2_6$.)

This suggests that I ought to be able to generalize my function `gen_sample`. Generalizing to any number of groups needs no extra work: the length of the input `n` determines the number of groups, and the values in `n` determine the size of each of those groups. 

The interesting generalization is the distribution to sample from. The first parameter of the  functions `rnorm`, `rchisq` etc. is always the number of random values to generate, but the remaining parameters are different for each distribution. This suggests that my generalized random sample generator ought to have the name of the random sampling function as input, followed by some means to allow any other inputs needed by that sampling function.


# Generalizing

To generalize `gen_sample` to a new function `sample_groups`, we need to consider how to handle different distributions. The distribution itself is not the hard part; that can be specified by having the random sample generator function for the desired distribution as an input to the new function. The problem is that each distribution has different parameters, which need to be inputs to `sample_groups`.

The standard way of doing this kind of thing is to use the `...` input to a function. If I had just one group, it would go like this:

```{r}
sample_group <- function(n, dist, ...) {
  dist(n, ...)
}
```

and then

```{r}
sample_group(5, rnorm, mean = 10, sd = 3)
```

or 

```{r}
sample_group(6, rpois, lambda = 2)
```

The additional (named) inputs to `sample_group` are passed on unchanged to the random sample generator, to generate respectively normal random values with mean 10 and SD 3, and Poisson random values with mean 2.^[It may take a trip to the help files to find out what R calls these parameters.]

The random sample generators are vectorized, but the obvious thing for generating two samples from different distributions does not work:

```{r}
sample_group(c(6, 3), rnorm, mean = c(10, 20), sd = c(3, 2))
```

We appear to have one value from each distribution, not six from the first and three from the second. This, I *think*, is what the help files say will happen.

To allow for the stuff at the end of the call to be different, another way is to use a `list` to pass each distribution's parameters. This turns out to be what I do later.


# A bad approach

Let's suppose I ask the user to write the code to generate each sample as text (a vector of pieces of text, one for each sample). Here's how my example above would look:

```{r}
code <- c("rnorm(5, mean = 10, sd = 3)", 
          "rpois(6, lambda = 2)")
code
```

The problem is that this is text, not runnable code. One way to turn this into something useful is to `parse` it:

```{r}
pc <- parse(text = code[1])
pc
```
This has turned the text into an `expression`, something that can be evaluated, thus:

```{r}
eval(pc)
```

And now we have a strategy:

```{r}
tibble(code) %>% 
  rowwise() %>% 
  mutate(expr = list(parse(text = code))) %>%
  mutate(z = list(eval(expr))) %>% 
  unnest(z)
```


So now we have the ingredients for a version of `sample_groups` based on the user writing the random-sampling code for us. I added one extra thing: lettering the groups, since they otherwise have bad names:

```{r}
sample_groups <- function(code) {
  n_pop <- length(code)
  tibble(code, gp = letters[1:n_pop]) %>% 
    rowwise() %>% 
    mutate(expr = list(parse(text = code))) %>%
    mutate(z = list(eval(expr))) %>% 
    unnest(z) %>% 
    select(gp, z)
}
```

and to test:

```{r}
d <- sample_groups(code)
d
```


Using the `eval(parse(text = something))` idea is not (apparently) very well regarded.^[There seems to be a recurring theme on Stack Overflow that if `eval(parse())` is the answer, you are asking the wrong question.] One immediate problem is that the user could put any code at all (that evaluates into a vector of numbers) into the input `code`, which seems less than secure:

```{r}
code <- c("1:3", "mtcars$mpg")
sample_groups(code)
```

# A better way

I want to get back to the user inputting the desired random sample generators as functions, and then running those functions on the rest of the input.
This is what `do.call` does:

```{r}
do.call(rnorm, list(n = 5, mean = 10, sd = 3))
do.call(rpois, list(n = 6, lambda = 2))
```

Having realized (i) that `do.call` is what I wanted, and (ii) that the input parameters to the functions need to be in a `list`, I packaged up those distribution parameters into a `list` of `list`s first. It is actually not necessary to make the list of distributions into a `list`, but it works if you do that too:

```{r}
dist <- list(rnorm, rpois)
pars <- list(list(n = 5, mean = 10, sd = 3), 
             list(n = 6, lambda = 2))
d <- tibble(dist = dist, pars = pars)
d
```

and then we put the `do.call` in a rowwise mutate, wrapping the whole thing in a list to make a list-column:

```{r}
d %>% 
  rowwise() %>% 
  mutate(z = list(do.call(dist, pars))) %>% 
  unnest(z)
```


it works!

And so, we can now make our function:

```{r}
sample_groups <- function(dist, pars) {
  nr <- length(pars)
  tibble(dist = dist, pars = pars, gp = letters[1:nr]) %>% 
    rowwise() %>% 
    mutate(z = list(do.call(dist, pars))) %>% 
    unnest(z) %>% 
    select(gp, z)
}
```

and to test

```{r}
dists <- list(rnorm, rpois)
pars <- list(list(n = 5, mean = 10, sd = 3), list(n = 6, lambda = 2))
sample_groups(dists, pars)
```

The only weirdness is that the user has to specify a list of lists for the parameters (because `do.call` needs a list for inputs to its function). But it definitely works.

One shortcut is that if you want all the samples to be from the same distribution, you specify only one thing in the input that I called `dists`:^[This is where the `list()` is important: there is no problem having a list-column of functions, but you cannot have a column which is just a function.]

```{r}
dist <- list(rnorm)
new_pars <- list(list(n = 3, mean = 5, sd = 1),
                 list(n = 2, mean = 10, sd = 2),
                 list(n = 4, mean = 20, sd = 3))
sample_groups(dist, pars = new_pars)
```

